How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-01-08 doc/paper.tex
more citations and improving intro

  The effects of the historical covariates are extremely small and vary considerably over time. For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place. The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e. class-imbalance) and the extremely stochastic nature of species survival. Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.  This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

@@

  Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis. We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.

@@

Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.

@@

\section{Data Specifications}

@@

\section{Model Specifications}

@@

\section{Model Parameter Estimation}

@@

\section{Model adequacy}

@@

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.

@@

\section{Model adequacy}

@@

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.

We can also compare the by-taxon in-sample performances of the four models.

Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.

\section{Cross-validation}

The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig. \ref{fig:fold_auc}). The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect. This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
% ROC OOS estimate

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

@@

We can also compare expected out-of-sample performance based on taxonomic group

Finally, expected out-of-sample performance is compared between taxonomic groups over time.

\section{Cross-validation}

Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data CITATION. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data CITATION. Additionally, multimodality increases with model complexity (Fig. \ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

>>>>

  The effects of the historical covariates are extremely small and vary considerably over time. For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place. The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e. class-imbalance) and the extremely stochastic nature of species survival. We find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.  This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

@@

  Ultimately, our results demonstrate that, given two random species, there is an approximately 77\% probability of correctly ranking their relative risks of extinction. We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis. We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

@@

Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.

@@

\section{Materials and Methods}

\subsection{Data Specifications}

@@

\subsection{Model Specifications}

@@

\subsection{Model Parameter Estimation}

@@

\subsection{Model adequacy}

@@

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.

@@

\section{Results}

\subsection{Model adequacy}

@@

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

@@

\subsection{Cross-validation}

Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}. Additionally, multimodality increases with model complexity (Fig. \ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.
